% !TEX root = main.tex

% --- ABSTRACT ---
\newpage
\markboth{ABSTRACT}{}
\begin{abstract}
\noindent
Agentic Artificial Intelligence refers to AI systems capable of performing tasks through autonomous or semi-autonomous agents that can reason, use tools, interact with external systems, and collaborate within structured workflows. In the financial services industry, the adoption of such systems requires not only advanced AI capabilities, but also strong guarantees in terms of scalability, control, auditability, and operational reliability.

\vspace{1em}

\noindent
This internship report presents the design and implementation of an \textbf{AI Agentic Hub Platform} aimed at building, deploying, and operating intelligent applications at scale. The proposed platform is based on a modular multi-agent architecture in which agents are generated from declarative configurations, deployed as independent containerized services, and coordinated through an orchestration layer. This approach makes it possible to create specialized agents without modifying their source code, by relying on configurable prompts, tools, model parameters, and input/output schemas.

\vspace{1em}

\noindent
The work focuses on several key aspects of enterprise AI deployment, including agent generation, workflow orchestration, schema-based communication, prompt and schema versioning using MLflow, Docker-based containerization, and observability. The objective is to provide a controlled and extensible blueprint for developing financial AI applications while reducing fragmentation, improving governance, and ensuring traceability across the full lifecycle of intelligent systems.

\vspace{1em}

\noindent
\textbf{Keywords:} Agentic AI, Multi-Agent Systems, AI Orchestration, Docker, MLflow, Financial Services, Schema Validation, Observability.
\end{abstract}

% --- SUMMARY ---
\newpage
\section*{Summary}
\markboth{SUMMARY}{}

This internship project focuses on the design and implementation of \textbf{VIP -- Value Intelligent Platform}, an \textbf{Enterprise-Grade Agentic AI Hub} aimed at building, deploying, and operating intelligent applications at scale.

\begin{itemize}
    \item \textbf{Dynamic Agent Generation:} Developed a configurable Agent Factory capable of creating specialized AI agents from declarative configurations, without modifying the source code.
    
    \item \textbf{Modular and Containerized Architecture:} Designed agents as independent FastAPI services deployed in isolated Docker containers, ensuring scalability, separation of concerns, and easier maintenance.
    
    \item \textbf{Centralized Governance and Versioning:} Integrated MLflow to manage prompts, schemas, and configuration artifacts, enabling traceability, version control, and a ``Zero Code Change'' approach.
    
    \item \textbf{Intelligent Orchestration:} Implemented an orchestration layer supported by an adapter to coordinate multi-agent workflows and manage schema transformations between agents.
\end{itemize}

\textbf{Conclusion:} The platform provides a scalable and reusable blueprint for enterprise AI applications, allowing organizations to transform complex business processes into modular, auditable, and intelligent agent-based workflows.

% --- ACKNOWLEDGMENTS ---
\newpage
\section*{Acknowledgments}
\markboth{ACKNOWLEDGMENTS}{}

I would first like to express my sincere gratitude to the entire team at \textbf{Value Digital Services} for welcoming me into such a professional, collaborative, and innovative environment. The experience gained throughout this internship has been both enriching and inspiring, and I truly appreciated the support and positive atmosphere within the company.
\vspace{1em}

I am deeply grateful to my professional supervisor, Mr. \textbf{Ahmed REBAI}, for his guidance, technical expertise, and continuous support throughout this project. His vision, advice, and constructive feedback played a major role in the successful realization of the \textit{VIP -- Value Intelligent Platform}.
\vspace{1em}

I would also like to sincerely thank my academic supervisor, Ms. \textbf{Jihen JERBI}, for her valuable supervision, encouragement, and academic guidance. Her support helped ensure the quality and rigor of this work throughout the internship period.
\vspace{1em}

On a personal note, I would like to express my deepest gratitude to my parents, \textbf{Hayet} and \textbf{Ali}, for their unconditional love, sacrifices, and constant encouragement throughout my studies. Their trust and support have always been my greatest source of motivation.
\vspace{1em}

I would also like to warmly thank my sister \textbf{Dhouha}, as well as my brothers \textbf{Borhen} and \textbf{Louey}, for their continuous encouragement, patience, and support during both the challenging and rewarding moments of this journey.
\vspace{1em}

Finally, I would like to thank everyone who contributed, directly or indirectly, to the success of this internship experience and to the completion of this project. This experience has been a major step in both my academic and professional journey, and I am truly grateful for the knowledge and experience I have gained.

% --- TABLE OF CONTENTS ---
\clearpage
\phantomsection
\thispagestyle{frontmatterplain}
\markboth{TABLE OF CONTENTS}{}
\tableofcontents

% --- LIST OF FIGURES ---
\clearpage
\phantomsection
\thispagestyle{frontmatterplain}
\markboth{LIST OF FIGURES}{}
\listoffigures

% --- LIST OF TABLES ---
\clearpage
\phantomsection
\thispagestyle{frontmatterplain}
\markboth{LIST OF TABLES}{}
\listoftables

% --- GLOSSARY ---
\clearpage
\section*{Glossary}
\phantomsection
\thispagestyle{frontmatterplain}
\markboth{GLOSSARY}{}
\addcontentsline{toc}{section}{Glossary}

\vspace{1.8cm}

\begingroup
\renewcommand{\arraystretch}{1.55}
\noindent
\begin{tabular}{@{}p{3cm}p{10cm}@{}}
    \textbf{\textit{ADLC:}} & Agent Development Lifecycle \\
    \textbf{\textit{AI:}} & Artificial Intelligence \\
    \textbf{\textit{API:}} & Application Programming Interface \\
    \textbf{\textit{IBM:}} & International Business Machines \\
    \textbf{\textit{LLM:}} & Large Language Model \\
    \textbf{\textit{QA:}} & Quality Assurance \\
    \textbf{\textit{REST:}} & Representational State Transfer \\
    \textbf{\textit{UI:}} & User Interface \\
    \textbf{\textit{UX:}} & User Experience \\
    \textbf{\textit{VIP:}} & Value Intelligent Platform \\
\end{tabular}
\endgroup
